### 2014-08-25 standing-variation-paper.tex
adding more files
wrote part of discussion

References on parallel adapation, standing variation, etc.:
  coop,
  pennings and hermisson.
Note pennings and hermisson found standing deleterious variation was very important.

@@

Stick to constant-speed waves.
Note monty's empircal speeds paper.

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the haploid fitness
\gc{heterozygous fintess, with no reference to the fitness of the
  homozygote for the mutant alleles}.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

% Branching-process-sized point mass approx to local groups present at $t=0$
% -- depends on local population density large enough relative to migration (to be "branching")
% -- and migration not too large (since we approximate as a point mass).
% Find mean density of clusters of successful standing mutants. 
% Is greater-than-one size of clusters likely to make a difference?

@@

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendents of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendents are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}

@@

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Sharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd ,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd, Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd
  Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their prescence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to various compounds, notably
those present in fava beans.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ %
{\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
  the frequency of G6pd-deficiency alleles. Size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
  \url{http://www.malariajournal.com/content/12/1/418}. Countries with
  endemic malaria are colored yellow. Figure taken from \citep{Howes-g6pd-variants}.

@@

Thoughts
species coherence.
The Biological Reality of Species: Gene Flow, Selection, and Collective Evolution

>>>>

References on parallel adaptation, standing variation, etc.:
  coop,
  pennings and hermisson.
Note pennings and hermisson found standing deleterious variation was very important.

@@

Stick to constant-speed waves.
Note monty's empirical speeds paper.

@@

For concreteness, suppose that before time $t=0$,
the mutant type has fitness $1-s_d$ relative to the neutral type
(i.e.\ it produces on average $1-s_d$ times the number of offspring per generation),
and that after time $t=0$,
the mutant type has fitness $1+s_b$,
where $s_b>0$ will usually be assumed to be small, and $0<s_d<1$.
We assume that diploid fitness is additive, or at least that the
important early dynamics are determined by the haploid fitness
\gc{heterozygous fitness, with no reference to the fitness of the
  homozygote for the mutant alleles}.
The number of offspring has finite variance, which to keep the formulas simple we assume to be equal to one.
(If the mutation is recessive, XXX more complicated things happen.)
As for the other parameters,
suppose that each offspring of a neutral parent is of the mutant type with probability $\mu$,
and that the mean squared distance between parent and child is $\sigma^2$ (the {\em dispersal variance}).
The species occupies an area with mean density
$\rho$ haploid individuals (or chromosomes) per unit area.
\marginnote{XXX or do diploid? below is haploid.}

% Branching-process-sized point mass approx. to local groups present at $t=0$
% -- depends on local population density large enough relative to migration (to be "branching")
% -- and migration not too large (since we approximate as a point mass).
% Find mean density of clusters of successful standing mutants. 
% Is greater-than-one size of clusters likely to make a difference?

@@

Before $t=0$, on the other hand, the mutation is deleterious, so the
genetic descendants of each new mutation are (with high probability)
doomed (doomed I say!) to extinction,
but may persist for some time.
Since the times and locations of mutants before $t=0$ also well-approximated by a Poisson process with rate $\mu \rho$,
the locations of all mutant families extant at $t=0$ whose descendants are destined to fix locally is also,
by the Mapping theorem, a Poisson process with rate we define to be $\lambda_0$.
If we assume that the descendants of at most only a few members of any extant mutant family at $t=0$ will survive,
and that these progenitors are near to each other in space, we can then treat each such family as a single mutant,
equivalent to the mutants arising later.
(The approximation will be good if the log of the size of an extant families is small relative to the establishment time,
and the spatial distribution of the family is small relative to the spread between them.)
To find $\lambda_0$, consider a mutation that arose at time $-T<0$, and let $Z_s$ be the number of its descendants at time $s-T$.
At time $t=0$, there are $Z_T$ individuals present with the mutation,
and each has probability $p_s \approx 2s_b$ of establishing, approximately independently.
Therefore, the probability that some descendants of this mutation establish and fix locally is $1-(1-p_s)^{Z_T}$,
so defining $\zeta(u,t) = \E[u^{Z_t} | Z_0=1 ]$ to be the generating function of $Z_T$,
the rate is
\begin{align*}
    \mu \rho \int_0^\infty \left( 1- \zeta(1-p_s,T) \right) dT .
\end{align*}
Now for $s_b$ small, using $p_s \approx 2s_b$ and that $\E[Z_T]=(1-s_d)^T$,
we know that $\zeta(1-2s_b,T) \approx 1-2s_b \E[Z_T] = 1-2s_b (1-s_d)^T$,
resulting in the approximation
\begin{align}
    \lambda_0 &= \mu \rho \int_0^\infty 2s_b (1-s_d)^{T} dT \\
        &= \frac{ 2 \mu \rho s_b }{ -\log(1-s_d) } .
\end{align}

@@

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd ,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd, Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd
  Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ %
{\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.

@@

Confusing convergence for local adaptation. Mention skin pigmentation?

Our results are predicated on the idea that adaptive variants are
initially rare within populations, i.e. they are reasonably
deleterious before the environment switches. In constrast, even distant populations could have
a shared basis of adaptation if they adapt via common variation,
e.g. previously neutral (or nearly neutral) variation, shared among populations. 
Our ability to detect such shared convergent events is at the moment
quite limited. 
One way that selection pressures might be weak on alleles, is if they
are at only one of many loci contributing to variation in a trait. However, if that
case populations separated by a reasonable amount of drift may still
adapt via different genetic routes, as the constellation of alleles
best positioned to initially respond to positive selection, i.e. those at
intermediate frequency, will be somewhat different.

\paragraph{}
Our results touch on an old debate on the evolutionary coherence of
species. 
Mayr and others have argued that species are coherent \citep{} 
as they are united by shared gene flow. However, this arguement has
questioned by a number of authors based on relatively high levels of 
differentiation in many species \citep{}.
\citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
may not be high enough to prevent neutral differentiation, species
should be viewed as being united as levels of gene flow were still
high enough for selected alleles to spread across the entire species.
Our work on geographic convergent adaptation
\citep[see also][]{ralph2010parallel,RalphCoop:14} brings these conclusions partially into
question. Our results suggest that wide-spread species should often
repeatedly adapt in parallel rather than waiting for a single allele
to migrate across the range.

In support of this idea recent selective sweeps are often
geographically restricted \citep{}, but see XXX. This is likely in part due to the
relatively low occurance of species-wide selection pressures \citep{},
but as noted above even when we know of wide-spread selection
pressures (e.g. Malaria) the response is usually convergent across large enough
spatial scales. This view is potentially contradicted by the
introgression of adaptive alleles across species and sub-species
boundaries, suggesting that selected alleles do sometimes spread
despite low migration rates \citep{}. However, at least some of
these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
alleles.

% \gc{Needed?:} This view is potentially contradicted by the
% introgression of adaptive alleles across species and sub-species
% boundaries, suggesting that selected alleles do sometimes spread
% despute low migration rates \citep{}. However, at least some of
% these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
% alleles. Other cases of selective introgression
% are concurrent with populations spreading into new environments,
% and coming in contact with preadapted populations, potentially
% lessening the chance of convergent adaptation \citep{}.

If gene flow rarely rapidly spread selected alleles across
species range, how then do selected alleles ever become fixed within species? 
Repeated bouts of adaptation in particular genomic regions may act to slowly
push a subset of previously selected alleles to fixation across the
species range, through the spread of the genetic backgrounds on which they arise.  
However, it seems likely that this is a slow process compared to the
initial rapid spread of selected alleles.

One potential hypothesis is that many selected alleles only achieve
fixation, not through their own species-wide spread, but rather
through subsequent large-scale
changes in geographic range size induced by extirpation of the species
over part of its range. Such drops in range size may fix alleles, or radically change the
species-wide frequency, of alleles previously restricted to small portion of a species range. 
Range sizes are in constant in flux, with

Similarly the range fragmentation accompanying speciation may act to fix variants within newly formed
species that were polymorphic across ancestral species ranges.

Note that this does not mean that:
Any change in neutral fixation rate.
Any change in

Many DMIs are locally restricted \citep{Cutter:12}.

species range evolution
splitting of species.
\citep{Rosenblum:12}

Futuyma has argued that much of the differentiation within species is \citep{Futuyma:10,FUTUYMA:87}
